---
title: "How accelerometers data can tell us how people exercise"
author: "Nathalie Descusse-Brown"
date: "April 02, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary
This paper makes use of data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. These participants were asked to perform barbell lifts correctly and incorrectly in 5 different ways. The aim of this paper is to use the data to predict the manner in which the participants  did the exercise. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).  

## Data Processing

```{r dataprocessing, cache=TRUE}
## load all data necessary to the assignment  
library(tidyverse)

if (!file.exists("pml-training.csv")) {
  fileUrl1 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(fileUrl1,destfile="pml-training.csv")
fileUrl2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(fileUrl2,destfile="pml-testing.csv")
}
train <- read_csv("pml-training.csv")
test <- read_csv("pml-testing.csv")

#summary(train)
```


## Exploratory Analysis

```{r exploratoryanalysis}
library(plyr)
library(Hmisc)


tidytrain <- data.frame(stringsAsFactors=FALSE)

j <- 1
for (i in 1:dim(train)[2])
        {
                if (((count(is.na(train[,i]))[[1,2]]>11773) & (count(train[,i]=="")[[1,2]]>7849) & (count(train[,i]=="#DIV/0!")[[1,2]]>7849)) & length(tidytrain)==0)
                {
                        tidytrain <- train[,i]
                        firstname <- colnames(train)[i]
                        j <- j+1
                }
                else if (((count(is.na(train[,i]))[[1,2]]>11773) & (count(train[,i]=="")[[1,2]]>7849) & (count(train[,i]=="#DIV/0!")[[1,2]]>7849)) & length(tidytrain)>0)
                {
                        tidytrain <- cbind(tidytrain,train[,i])
                        colnames(tidytrain)[j] <- colnames(train)[i]
                        j <- j+1
                }
        }
        

dim(tidytrain)
names(tidytrain)
```

We know from the data source that the data is classified as follows:
- Class A: performed exactly according to the speci???cation
- Class B: throwing the elbows to the front 
- Class C: lifting the dumbbell only halfway 
- Class D: lowering the dumbbell only halfway 
- Class E: throwing the hips to the front.

Because of the classification and the location of the measuring devices we make the following assumptions:
- Class B should be associated with a negative gyros_dumbbell_z as the dumbell is pushed to the shoulder when the elbow goes to the front.
- Class C should be associated with smaller gyros_dumbbell_y and gyros_forearm_y compared to Class A and Class B.
- Class D should be associated with similar amplitude gyros_dumbbell_y and gyros_forearm_y compared to Class C but shifted along the y axis.
- Class E should be associated with larger gyros_belt_z compared to other classes.


```{r ggpairexploratory, out.width = '1000px', dpi=300}
# library(ggplot2)
# library(GGally)
# totalcorr <- ggcorr(tidytrain, palette = "RdBu", label = TRUE)
# summary(totalcorr)

library(gridExtra) 
# g = ggplot(tidytrain,aes(y=roll_belt,x=classe,colour=user_name))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5)
# g2 = ggplot(tidytrain,aes(y=pitch_belt,x=classe,colour=user_name))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# g3 = ggplot(tidytrain,aes(y=yaw_belt,x=classe,colour=user_name))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# g4 = ggplot(tidytrain,aes(y=total_accel_belt,x=classe,colour=user_name))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# 
# grid.arrange(g, g2,g3,g4,nrow=2, ncol=2)
# 
# h = ggplot(tidytrain,aes(y=roll_forearm,x=classe,colour=classe))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5)
# h2 = ggplot(tidytrain,aes(y=pitch_forearm,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# h3 = ggplot(tidytrain,aes(y=yaw_forearm,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# h4 = ggplot(tidytrain,aes(y=total_accel_forearm,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# 
# grid.arrange(h, h2,h3,h4,nrow=2, ncol=2)
# 
# k = ggplot(tidytrain,aes(y=gyros_belt_x,x=classe,colour=classe))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5)
# k2 = ggplot(tidytrain,aes(y=gyros_belt_y,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# k3 = ggplot(tidytrain,aes(y=gyros_belt_z,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# 
# grid.arrange(k, k2,k3,nrow=2, ncol=2)
# 
# l = ggplot(tidytrain,aes(y=accel_belt_x,x=classe,colour=classe))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5)
# l2 = ggplot(tidytrain,aes(y=accel_belt_y,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# l3 = ggplot(tidytrain,aes(y=accel_belt_z,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#        geom_point(size=0.5)
# 
# grid.arrange(l, l2,l3,nrow=2, ncol=2)


o = ggplot(tidytrain,aes(y=yaw_belt,x=roll_belt,colour=classe))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5)
o

p = ggplot(tidytrain,aes(y=gyros_dumbbell_x,x=classe,colour=classe))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5) + ylim(-2,2)
p2 = ggplot(tidytrain,aes(y=gyros_dumbbell_y,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
        geom_point(size=0.5) + ylim(-2,2)
p3 = ggplot(tidytrain,aes(y=gyros_dumbbell_z,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
        geom_point(size=0.5) + ylim(-2,2)

grid.arrange(p,p2,p3,nrow=2, ncol=2)

# q = ggplot(tidytrain,aes(y=gyros_belt_x,x=classe,colour=classe))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5)
# q2 = ggplot(tidytrain,aes(y=gyros_belt_y,x=classe,colour=classe))+geom_point(colour="pink",size=0.5)+
#         geom_point(size=0.5)
q3 = ggplot(tidytrain,aes(y=gyros_belt_z,x=gyros_forearm_z,colour=classe))+geom_point(colour="pink",size=0.5)+
        geom_point(size=0.5) + xlim(-2,2)

q3

s = ggplot(tidytrain,aes(y=gyros_forearm_x,x=gyros_arm_x,colour=classe))+geom_point(colour="grey50",size=0.5)+geom_point(size=0.5) + ylim(-5,5)
# s2 = ggplot(tidytrain,aes(y=gyros_forearm_y,x=gyros_arm_y,colour=classe))+geom_point(colour="pink",size=0.5)+
#         geom_point(size=0.5)
# s3 = ggplot(tidytrain,aes(y=gyros_forearm_z,x=gyros_arm_z,colour=classe))+geom_point(colour="pink",size=0.5)+
#         geom_point(size=0.5)

s
# 
# grid.arrange(s,s2,s3,nrow=2, ncol=2)


tidytrainnum <- tidytrain[,sapply(tidytrain,is.numeric)]
MMMmat<-as.matrix(tidytrainnum)

M <- rcorr(MMMmat,type="pearson")

for (i in 1:dim(M$r)[1])
{
   for (j in 1:i)
   {
     if (!is.na(M$P[i,j]) & M$r[i,j]>0.8 & i!=j & M$P[i,j] < 0.05 )
     {
       print(paste(rownames(M$r)[i], "-" , colnames(M$r)[j], ": ", M$r[i,j], ": ", M$P[i,j]))
     }
   }
}

```


## Model Selection and Results

```{r prediction}
library(caret)
library(rattle)
library(MASS)
library(parallel)
library(doParallel)

#below taken from https://github.com/lgreski/datasciencectacontent/blob/master/markdown/pml-randomForestPerformance.md
cluster <- makeCluster(detectCores() - 1) 
registerDoParallel(cluster)
fitControl <- trainControl(method = "cv",
                           number = 5,
                           allowParallel = TRUE)


modelFit <- train(classe ~ .,method="rf",data=tidytrain,trControl = fitControl)

stopCluster(cluster)
registerDoSEQ()
# modelFitoptim <- step(modelFit)
# modelFit$finalModel
modelFit
modelFit$resample
confusionMatrix.train(modelFit)
```


## Discussion